---
title: "Principal Component Analysis"
output: html_document
date: "2022-12-09"
---

# Housekeeping

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('/Users/lena/documents/R/PCA')
library(tidyverse)
library(nnet)
library(yardstick)
library(devtools)
library(ggbiplot)
library(psych)
library(MVN)
```

# Load and prepare the data

```{r}
df_wine <- data.table::fread("wine.data")
colnames(df_wine) <- c("class", 
                       "alcohol",
                       "malic_acid", 
                       "ash",
                       "alca_ash",
                       "magnesium",
                       "total_phenols",
                       "flav",
                       "nonflav_phenols",
                       "proan",
                       "col_intens",
                       "hue",
                       "protein",
                       "proline")
```

# A first glance

```{r}

df_wine %>% 
  mutate(class = as.factor(class)) %>%  
  ggplot(aes(malic_acid, ash, color = class)) +
  geom_point()

df_wine %>% 
  mutate(class = as.factor(class)) %>%  
  ggplot(aes(protein, flav, color = class)) +
  geom_point()

```

# Barlett's sphericity test

$H_0: P = I \text{ vs. }H_1: P \neq I$

```{r}
z_score <- function(x){
  (x-mean(x))/sd(x)
}

data_testing <- df_wine %>% 
  select(-class)
data_testing <- data.frame(apply(data_testing, 2, z_score))

cor_testing <- cor(data_testing)
cortest.bartlett(cor_testing, n = nrow(data_testing))
```

We reject $H_0$. 
Are they mvn distributed?

# Test for multivariate normal distribution: Royston's H test
$H_0$: data is multivariate normally distributed vs. $H_1$ not $H_0$

```{r}
mvn_test <- mvn(data_testing, mvnTest = "royston")
mvn_test$multivariateNormality
```
We reject $H_0$ of multivariate normality.

# Split into test and train and define x and y

This is preparation for the classification later

```{r}
set.seed(48)
df_wine <- df_wine %>% 
  mutate(id = sample(x = c(0, 1), size = nrow(.), replace = TRUE, prob = c(.7, .3)))

train <- df_wine %>% 
  filter(id == 0) %>% 
  select(-id)
test <- df_wine %>% 
  filter(id == 1) %>% 
  select(-id)

train_x <- train[, c(2:14)]
train_y <- train[, 1]

test_x <- test[, c(2:14)]
test_y <- test[, 1]

```

# Standardise the x variables for PCA

```{r}
train_x <- data.frame(apply(train_x, 2, z_score))
test_x <- data.frame(apply(test_x, 2, z_score))
```

# Principal Component Analysis

```{r}
pca_cor <- cor(train_x)
pca_eigen <- eigen(pca_cor)
pca_eigenvec <- pca_eigen$vectors
pca_eigenval <- pca_eigen$values

pc1 <- t(pca_eigenvec[,1])%*%t(train_x) # compute first PC on training data
pc2 <- t(pca_eigenvec[,2])%*%t(train_x) # compute second PC on training data
pc3 <- t(pca_eigenvec[,3])%*%t(train_x) # compute third PC on training data
pc4 <- t(pca_eigenvec[,4])%*%t(train_x) # compute fourth PC on training data

pca_eigenval <- data.frame(pca_eigenval) %>%
  mutate(explained_var = pca_eigenval/sum(pca_eigenval)) %>% 
  mutate(cum_var = cumsum(explained_var))
pca_eigenval
```

# Visualisation: Number of principal components
Slides

# Visualisation: Interpretation of the principal components 
Slides


# Classification with logit regression

```{r}
train$PC1 <- t(pc1)
train$PC2 <- t(pc2)

logistic_classification <- multinom(class ~ PC1 + PC2,
                               data = train)

```

# Predict on test data

```{r}
pc1_test <- t(pca_eigenvec[,1])%*%t(test_x)
pc2_test <- t(pca_eigenvec[,2])%*%t(test_x) 
test_x$PC1 <- as.numeric(t(pc1_test))
test_x$PC2 <- as.numeric(t(pc2_test))

logit_predict <- predict(logistic_classification, test_x)
logit_predict

```

# Model Evaluation

```{r}
validation <- cbind(logit_predict, test_y)
validation <- data.frame(validation) %>% 
  mutate(class = as.factor(class)) %>% 
  mutate(compare = if_else(logit_predict == class, TRUE, FALSE))
validation

accuracy <- (sum(validation$compare))/nrow(validation)
accuracy

conf_matrix <- conf_mat(validation, class, logit_predict)

conf_matrix %>% 
  autoplot(type = "heatmap") +
  scale_fill_gradient(low="#D6EAF8",high = "#2E86C1") +
  theme(legend.position = "right") +
  labs(fill = "Abs. freq")

```